Long term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-06-11 article/article.Rmd
Commit 2024-06-11 13:11

According to this study, in the period 1997 -- 2005 the mean AOD at $340\,\text{nm}$
$0.403$ with a trend of $-3.8\pm0.93\,\%$ per year, corresponding to a change of
$0.0153$ per year.  Using an Ångström coefficient $\alpha = 1.6$, this translates to
a change in the Ångström coefficient $\beta=0.00272$ per year (or $\beta=0.084$ in
1997 and $\beta=0.059$ in 2005).  Simulations with uvspec for the above Ångström
coefficients, with WC of
$`r round(BR[year==1997,pw_avg_mm],1)`\,\text{mm}$ and 
$`r round(BR[year==2005,pw_avg_mm],1)`\,\text{mm}$
for 1997 and 2005 respectively and for a SZA of $55^\circ$ reveal trend of
$+0.21\,\%$ per year in $\text{GHI}_\text{ref}$. 
The SZA of $55^\circ$ was chosen as representative of all days in the year in order
to get a rough estimate of the annually averaged change in clear sky irradiance.  For
the period 2005 -- 2023 we used the mean monthly values of AOD and WC from AERONET in
a similar simulation scheme to calculate the monthly mean of clear sky irradiance,
and finally the trend of $+0.14\,\%$ per year. We applied these two long term changes
(see Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
$\text{GHI}_\text{ref}$, in order to create a more realistic representation of the
clear-sky irradiance for the whole period of study.

@@

First, we evaluated the performance of the modelled $\text{GHI}_\text{ref}$ in
relation to the measured GHI for different levels of atmospheric clearness, by using
in the RTM the monthly climatological AOD and WC, less their respective standard
deviations.
<!-- $\tau_{\text{cs}} = \tau_{500\text{nm}} - 1\sigma$ and $w_{h\text{cs}} = w_h - 1\sigma$
($\text{GHI}_\text{CSm}$). -->
These values represent typical atmosphere in Thessaloniki, with lower than average
load of aerosols and humidity, which are the main factor that attenuate the GHI,
excluding clouds.  With this approach the simulated $\text{GHI}_\text{ref}$ should be
generally greater than the measured GHI when aerosols are more abundant.  To
compensate for this, we increased the $\text{GHI}_\text{ref}$ by 
($`r 100*(C4_cs_ref_ratio-1)`\,\%$)
with an additional constant offset of
$`r C4_GLB_diff_THRES`\,\text{W}/\text{m}^2$,
as described in Equation\nobreakspace\ref{eq:CE4}.
\begin{equation}
\text{CE} : \text{E}_\text{i} > `r C4_GLB_diff_THRES` + `r C4_cs_ref_ratio` \cdot \text{E}_\text{CSm,i} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation}
where: $\text{E}$ the measured irradiance, $\text{E}_\text{CSm}$ the selected
modelled clear sky irradiance, and $i$ each of the one-minute observation.  This is
the criterion of our CE identification.

@@

<!--
- can follow for clear days
- this selection has some subjectivity
- the most important are the thresholds
- model have to be uniform on clear days
- the level is adjusted by the threshold 
- the effect of the sza uniform
-->

The definition of the CE events with this method has a degree of subjectivity, since
the actual clear sky irradiance, is not known and can only be approximated.
However, this method is capable in detecting all major CE events.  Where, some CE
events, with very low OIR may be not detected, these are few with small
over-irradiance and it is unlikely that will effect significantly our results.

<!-- extreme cases criterion -->
A sub-category  of the CE events that is often discussed in the relevant bibliography
[@Cordero2023; @Martins2022; @Yordanov2015], are the extreme cloud enhancement (ECE)
events. These are cases of CE where the measured intensity of the irradiance, exceeds
the TSI at the top of the atmosphere:
<!--and detected according to the Equation\nobreakspace{}\ref{eq:ECE}.-->
\begin{equation}
\text{ECE}: \text{GHI}_\text{i} > \cos(\theta) \times E_{i\odot} / r_{i}^2
\label{eq:ECE}
\end{equation}
where: $\theta$ the solar zenith angle, $E_{\odot}$ the solar constant adjusted for the actual Sun -- Earth distance, and $i$ each of the one-minute observation.

An example of CE identification for a selected day is given, in the
Figure\nobreakspace{}\ref{fig:example-day}, where the daily course of the clear sky
reference irradiance and the CE and ECE identification thresholds are shown along
with the actual GHI measurements.  In addition, we provide an example scatter plot
between the measured and the modeled clear-sky irradiance for one year, where the CE
and ECE events are clearly grouped above the threshold of irradiance
(Figure\nobreakspace{}\ref{fig:example-year}).

@@

Following the application of the above discussed methodology to the entire dataset 
(`r round(ST_total$wattGLB.N / 10^6,1)` million one-minute GHI measurements),
$`r round(100*ST_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ were identified as CE events and
$`r round(100*ST_extreme_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ as ECE events.
The highest recorded GHI due to CE was
$`r round(ST_daily[which.max(wattGLB.max), wattGLB.max],1)`\,\text{W}/\text{m}^2$
on `r strftime(ST_daily[which.max(wattGLB.max), Date], "%d\\ %B %Y")`, corresponding
to OIR of $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 1)`\,\text{W}/\text{m}^2$.
<!-- The absolute strongest CE event had an OIR of 
<!-- $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 2)`\,\text{W}/\text{m}^2$ on
<!-- `r strftime(ST_daily[which.max(GLB_diff.max), Date], "%d\\ %B %Y")`. -->
The stronger CE event of
$`r round(100*ST_daily[which.max(GLB_ench.max), GLB_ench.max],1)`\,\%$ above the 
clear sky threshold was observed on 
`r strftime(ST_daily[which.max(GLB_ench.max), Date], "%d\\ %B %Y")`.
In the following sections we are discussing the long-term trends and variability of the CE events as well as of the corresponding OIR and excess irradiation .

@@

We computed the daily trend of the mean OIR of CE <!--
(Figure\nobreakspace{}\ref{fig:CEmeanDaily}), --> using a first-order autoregressive
model with a lag of one day, with the 'maximum likelihood' fitting method
[@Gardner1980; @Jones1980] by implementing the function 'arima' from the library
'stats' of the R programming language [@RCT2023]. The trends are reported together
with the $2\sigma$ error.  We observed a negligible change of
$`r round(dailyCE[var == "GLB_diff.mean", Tmod_Estimate], 3)`\pm
`r round(2 * dailyCE[var == "GLB_diff.mean", Tmod_Std..Error], 3)`\,\text{W}/\text{m}^2/\text{year}$
on the mean OIR.
<!-- ```{r CEmeanDaily, echo=FALSE, fig.cap="Daily mean values and trend of the CE over irradiance."} -->
<!-- knitr::include_graphics("../images/P_daily_trend-1.png") -->
<!-- ``` -->
Although, the previous result is closer to the raw data, we preferred to present the
annual statistics, that give a more clear picture about the long term CE trends.
Hence, we calculated the annual values from the one-minute measurements.  Despite the
very weak statistical significance of the mean OIR trend, the annual number of CE
occurrences shows a steady increase of
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$
(Figure\nobreakspace{}\ref{fig:P-energy-N}).
Subsequently, the mean annual energy of the CE events is increasing with a rate of
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kj}/\text{year}$
(Figure\ \ref{fig:P-energy-sum}), witch follows the trend of the number of CE
occurrences.  We have to note that the energy related to the CE events can not be
directly linked with the total energy balance of the atmosphere. The net solar
radiation of the region is not increased, but rather redistributed through the CE
events.  This also is depicted by the "instantaneous" values of irradiance (average
over one minute), that can exceed the equivalent clear sky irradiance to a
considerable level.

@@

<!-- ```{r CEmeanDaily2, echo=FALSE, fig.cap="Trend of the daily excess energy due to CE"} -->
<!-- knitr::include_graphics("../images/daily-12.png") -->
<!-- ``` -->

<!-- ```{r echo=FALSE, fig.cap="Treds of the mean over irradiance per CE"} -->
<!-- knitr::include_graphics("../images/energy-3.png") -->
<!-- ``` -->

<!-- ```{r P-energy-mean, echo=FALSE, fig.cap="Trends of the mean OIR per CE."} -->
<!-- knitr::include_graphics("../images/P-energy-3.png") -->
<!-- ``` -->

<!-- ```{r echo=FALSE, fig.cap="Trend of number of CE cases"} -->
<!-- knitr::include_graphics("../images/energy-2.png") -->
<!-- ``` -->

<!-- ```{r P-energy-N, echo=FALSE, fig.cap="Trend of yearly cloud enhancement occurrences."} -->
<!-- knitr::include_graphics("../images/P-energy-2.png") -->
<!-- ``` -->

<!-- ```{r echo=FALSE, fig.cap="Trend of the yearly excess energy due to CE over irradiance"} -->
<!-- knitr::include_graphics("../images/energy-1.png") -->
<!-- ``` -->

<!-- ```{r P-energy-sum, echo=FALSE, fig.cap="Trend of the yearly excess energy due to CE over irradiance"} -->
<!-- knitr::include_graphics("../images/P-energy-1.png") -->
<!-- ``` -->

<!-- ```{r P-energy-median, echo=FALSE, fig.cap="Trend of the yearly median over irradiance due to CE over irradiance"} -->
<!-- knitr::include_graphics("../images/P_energy-8.png") -->
<!-- ``` -->

@@

\begin{figure}[h!]
        {\centering 
            \subfloat[All skies.\label{fig:cusumnotrendmonthly-1}]
                {\includegraphics[width=\linewidth]{../images/P-energy-3} }\\
            \subfloat[Clear skies.\label{fig:cusumnotrendmonthly-2}]
                {\includegraphics[width=\linewidth]{../images/P-energy-2} }\\
            \subfloat[Cloudy skies.\label{fig:cusumnotrendmonthly-3}]
                {\includegraphics[width=\linewidth]{../images/P-energy-1} }
        }
    \caption{Time series for the period 1992 -- 2023 of (a) the yearly CE number of occurrences, (b) the yearly mean OIR and (c) the yearly excess irradiation. The black lines represent the linear trends on the yearly data.}\label{fig:cusumnotrendmonthly}
\end{figure}

@@

# Appendix {.unnumbered}

<!-- # References {.unnumbered} -->

>>>>

According to this study, in the period 1997 -- 2005 the mean AOD at $340\,\text{nm}$
$0.403$ with a trend of $-3.8\pm0.93\,\%$ per year, corresponding to a change of
$0.0153$ per year.  Using an Ångström coefficient $\alpha = 1.6$, this translates to
a change in the Ångström coefficient $\beta=0.00272$ per year (or $\beta=0.084$ in
1997 and $\beta=0.059$ in 2005).  Simulations with uvspec for the above Ångström
coefficients, with WC of
$`r round(BR[year==1997,pw_avg_mm],1)`\,\text{mm}$ and 
$`r round(BR[year==2005,pw_avg_mm],1)`\,\text{mm}$
for 1997 and 2005 respectively and for a SZA of $55^\circ$ reveal a trend of
$+0.21\,\%$ per year in $\text{GHI}_\text{ref}$. 
The SZA of $55^\circ$ was chosen as representative of all days in the year in order
to get a rough estimate of the annually averaged change in clear sky irradiance.  For
the period 2005 -- 2023 we used the mean monthly values of AOD and WC from AERONET in
a similar simulation scheme to calculate the monthly mean of clear sky irradiance,
and finally the trend of $+0.14\,\%$ per year. We applied these two long term changes
(see Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
$\text{GHI}_\text{ref}$, in order to create a more realistic representation of the
clear-sky irradiance for the whole period of study.

@@

First, we evaluated the performance of the modelled $\text{GHI}_\text{ref}$ in
relation to the measured GHI for different levels of atmospheric clearness, by using
in the RTM the monthly climatological AOD and WC, less their respective standard
deviations.
<!-- $\tau_{\text{cs}} = \tau_{500\text{nm}} - 1\sigma$ and $w_{h\text{cs}} = w_h - 1\sigma$
($\text{GHI}_\text{CSm}$). -->
These values represent typical atmosphere in Thessaloniki with lower than average
load of aerosols and humidity, which are the main factor that attenuate the GHI,
excluding clouds.  With this approach the simulated $\text{GHI}_\text{ref}$ should be
generally greater than the measured GHI when aerosols are more abundant.  To
compensate for this, we increased the $\text{GHI}_\text{ref}$ by 
($`r 100*(C4_cs_ref_ratio-1)`\,\%$)
with an additional constant offset of
$`r C4_GLB_diff_THRES`\,\text{W}/\text{m}^2$,
as described in Equation\nobreakspace\ref{eq:CE4}.
\begin{equation}
\text{CE} : \text{E}_\text{i} > `r C4_GLB_diff_THRES` + `r C4_cs_ref_ratio` \cdot \text{E}_\text{CSm,i} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation}
where: $\text{E}$ the measured irradiance, $\text{E}_\text{CSm}$ the selected
modelled clear sky irradiance, and $i$ each of the one-minute observation.  This is
the criterion of our CE identification.

@@

<!--
- can follow for clear days
- the most important are the thresholds
- model have to be uniform on clear days
- the level is adjusted by the threshold 
- the effect of the sza uniform
-->

The definition of the CE events with this method has a degree of subjectivity, since
the actual clear sky irradiance is not known and can only be approximated.  However,
this method is capable in detecting all major CE events.  Where some CE events with
very low OIR may be not detected, these are few with small over-irradiance and it is
unlikely that will effect significantly our results.

<!-- extreme cases criterion -->
A sub-category  of the CE events that is often discussed in the relevant bibliography
[@Cordero2023; @Martins2022; @Yordanov2015], are the extreme cloud enhancement (ECE)
events. These are cases of CE where the measured intensity of the irradiance exceeds
the TSI at the top of the atmosphere:
<!--and detected according to the Equation\nobreakspace{}\ref{eq:ECE}.-->
\begin{equation}
\text{ECE}: \text{GHI}_\text{i} > \cos(\theta) \times E_{i\odot} / r_{i}^2
\label{eq:ECE}
\end{equation}
where: $\theta$ the solar zenith angle, $E_{\odot}$ the solar constant adjusted for the actual Sun -- Earth distance, and $i$ each of the one-minute observation.

An example of CE identification for a selected day is given in the
Figure\nobreakspace{}\ref{fig:example-day}, where the daily course of the clear sky
reference irradiance and the CE and ECE identification thresholds are shown along
with the actual GHI measurements.  In addition, we provide an example scatter plot
between the measured and the modeled clear-sky irradiance for one year, where the CE
and ECE events are clearly grouped above the threshold of irradiance
(Figure\nobreakspace{}\ref{fig:example-year}).

@@

Following the application of the above discussed methodology to the entire dataset 
(`r round(ST_total$wattGLB.N / 10^6,1)` million of one-minute GHI measurements),
$`r round(100*ST_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ were identified as CE events and
$`r round(100*ST_extreme_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ as ECE events.
The highest recorded GHI due to CE was
$`r round(ST_daily[which.max(wattGLB.max), wattGLB.max],1)`\,\text{W}/\text{m}^2$
on `r strftime(ST_daily[which.max(wattGLB.max), Date], "%d\\ %B %Y")`, corresponding
to OIR of $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 1)`\,\text{W}/\text{m}^2$.
<!-- The absolute strongest CE event had an OIR of 
<!-- $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 2)`\,\text{W}/\text{m}^2$ on
<!-- `r strftime(ST_daily[which.max(GLB_diff.max), Date], "%d\\ %B %Y")`. -->
The stronger CE event of
$`r round(100*ST_daily[which.max(GLB_ench.max), GLB_ench.max],1)`\,\%$ above the 
clear sky threshold was observed on 
`r strftime(ST_daily[which.max(GLB_ench.max), Date], "%d\\ %B %Y")`.
In the following sections we are discussing the long-term trends and variability of the CE events as well as of the corresponding OIR and excess irradiation .

@@

The main focus of this study is to investigate the time evolution of the CE events by
analyzing the GHI measurements at Thessaloniki. Cloud enhancements can be influenced
by different factors, such as the geometry, size and optical thickness of clouds,
their height in the atmosphere and on local weather regimes **!!!(ref)!!!**.  Some of
these factors are related to changes in climate; hence it would be reasonable to
expect capturing their contributions to the frequency of occurrence of CE events over
Thessaloniki, as well as to the average OIR and excess irradiation. 
The long-term trends were calculated using a first-order autoregressive model with the
'maximum likelihood' fitting method
[@Gardner1980; @Jones1980], by implementing the function 'arima' from the library
'stats' of the R programming language [@RCT2023]. All trends are reported together
with their $2\sigma$ error.

We observed a negligible change of
$`r round(dailyCE[var == "GLB_diff.mean", Tmod_Estimate], 3)`\pm
`r round(2 * dailyCE[var == "GLB_diff.mean", Tmod_Std..Error], 3)`\,\text{W}/\text{m}^2/\text{year}$
on the mean OIR.
<!-- ```{r CEmeanDaily, echo=FALSE, fig.cap="Daily mean values and trend of the CE over irradiance."} -->
<!-- knitr::include_graphics("../images/P_daily_trend-1.png") -->
<!-- ``` -->
Although, the previous result is closer to the raw data, we preferred to present the
annual statistics, that give a more clear picture about the long term CE trends.
Hence, we calculated the annual values from the one-minute measurements.  Despite the
very weak statistical significance of the mean OIR trend, the annual number of CE
occurrences shows a steady increase of
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$
(Figure\nobreakspace{}\ref{fig:P-energy-N}).
Subsequently, the mean annual energy of the CE events is increasing with a rate of
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kj}/\text{year}$
(Figure\ \ref{fig:P-energy-sum}), witch follows the trend of the number of CE
occurrences.  We have to note that the energy related to the CE events can not be
directly linked with the total energy balance of the atmosphere. The net solar
radiation of the region is not increased, but rather redistributed through the CE
events.  This also is depicted by the "instantaneous" values of irradiance (average
over one minute), that can exceed the equivalent clear sky irradiance to a
considerable level.

@@

<!-- ```{r CEmeanDaily2, echo=FALSE, fig.cap="Trend of the daily excess energy due to CE"} -->
<!-- knitr::include_graphics("../images/daily-12.png") -->
<!-- ``` -->
<!-- ```{r echo=FALSE, fig.cap="Treds of the mean over irradiance per CE"} -->
<!-- knitr::include_graphics("../images/energy-3.png") -->
<!-- ``` -->
<!-- ```{r P-energy-mean, echo=FALSE, fig.cap="Trends of the mean OIR per CE."} -->
<!-- knitr::include_graphics("../images/P-energy-3.png") -->
<!-- ``` -->
<!-- ```{r echo=FALSE, fig.cap="Trend of number of CE cases"} -->
<!-- knitr::include_graphics("../images/energy-2.png") -->
<!-- ``` -->
<!-- ```{r P-energy-N, echo=FALSE, fig.cap="Trend of yearly cloud enhancement occurrences."} -->
<!-- knitr::include_graphics("../images/P-energy-2.png") -->
<!-- ``` -->
<!-- ```{r echo=FALSE, fig.cap="Trend of the yearly excess energy due to CE over irradiance"} -->
<!-- knitr::include_graphics("../images/energy-1.png") -->
<!-- ``` -->
<!-- ```{r P-energy-sum, echo=FALSE, fig.cap="Trend of the yearly excess energy due to CE over irradiance"} -->
<!-- knitr::include_graphics("../images/P-energy-1.png") -->
<!-- ``` -->
<!-- ```{r P-energy-median, echo=FALSE, fig.cap="Trend of the yearly median over irradiance due to CE over irradiance"} -->
<!-- knitr::include_graphics("../images/P_energy-8.png") -->
<!-- ``` -->

@@

\begin{figure}% [h!]
        {\centering 
            \subfloat[\label{fig:cusumnotrendmonthly-1}]
                {\includegraphics[width=\linewidth]{../images/P-energy-3} }\\
            \subfloat[\label{fig:P-energy-N}]
                {\includegraphics[width=\linewidth]{../images/P-energy-2} }\\
            \subfloat[\label{fig:cusumnotrendmonthly-3}]
                {\includegraphics[width=\linewidth]{../images/P-energy-1} }
        }
    \caption{Time series for the period 1992 -- 2023 of (a) the yearly CE number of occurrences, (b) the yearly mean OIR and (c) the yearly excess irradiation. The black lines represent the linear trends on the yearly data.}\label{fig:cusumnotrendmonthly}
\end{figure}

@@

### 2024-06-11 article/article.tex
Commit 2024-06-11 13:11

According to this study, in the period 1997 -- 2005 the mean AOD at
\(340\,\text{nm}\) \(0.403\) with a trend of \(-3.8\pm0.93\,\%\) per
year, corresponding to a change of \(0.0153\) per year. Using an
Ångström coefficient \(\alpha = 1.6\), this translates to a change in
the Ångström coefficient \(\beta=0.00272\) per year (or \(\beta=0.084\)
in 1997 and \(\beta=0.059\) in 2005). Simulations with uvspec for the
above Ångström coefficients, with WC of \(15.7\,\text{mm}\) and
\(15.4\,\text{mm}\) for 1997 and 2005 respectively and for a SZA of
\(55^\circ\) reveal trend of \(+0.21\,\%\) per year in
\(\text{GHI}_\text{ref}\). The SZA of \(55^\circ\) was chosen as
representative of all days in the year in order to get a rough estimate
of the annually averaged change in clear sky irradiance. For the period
2005 -- 2023 we used the mean monthly values of AOD and WC from AERONET
in a similar simulation scheme to calculate the monthly mean of clear
sky irradiance, and finally the trend of \(+0.14\,\%\) per year. We
applied these two long term changes (see
Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
\(\text{GHI}_\text{ref}\), in order to create a more realistic
representation of the clear-sky irradiance for the whole period of
study.

@@

First, we evaluated the performance of the modelled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmosphere in Thessaloniki, with lower
than average load of aerosols and humidity, which are the main factor
that attenuate the GHI, excluding clouds. With this approach the
simulated \(\text{GHI}_\text{ref}\) should be generally greater than the
measured GHI when aerosols are more abundant. To compensate for this, we
increased the \(\text{GHI}_\text{ref}\) by (\(4\,\%\)) with an
additional constant offset of \(15\,\text{W}/\text{m}^2\), as described
in Equation\nobreakspace\ref{eq:CE4}. \begin{equation}
\text{CE} : \text{E}_\text{i} > 15 + 1.04 \cdot \text{E}_\text{CSm,i} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation} where: \(\text{E}\) the measured irradiance,
\(\text{E}_\text{CSm}\) the selected modelled clear sky irradiance, and
\(i\) each of the one-minute observation. This is the criterion of our
CE identification.

@@

The definition of the CE events with this method has a degree of
subjectivity, since the actual clear sky irradiance, is not known and
can only be approximated. However, this method is capable in detecting
all major CE events. Where, some CE events, with very low OIR may be not
detected, these are few with small over-irradiance and it is unlikely
that will effect significantly our results.

A sub-category of the CE events that is often discussed in the relevant
bibliography \citep{Cordero2023, Martins2022, Yordanov2015}, are the
extreme cloud enhancement (ECE) events. These are cases of CE where the
measured intensity of the irradiance, exceeds the TSI at the top of the
atmosphere: \begin{equation}
\text{ECE}: \text{GHI}_\text{i} > \cos(\theta) \times E_{i\odot} / r_{i}^2
\label{eq:ECE}
\end{equation} where: \(\theta\) the solar zenith angle, \(E_{\odot}\)
the solar constant adjusted for the actual Sun -- Earth distance, and
\(i\) each of the one-minute observation.

An example of CE identification for a selected day is given, in the
Figure\nobreakspace{}\ref{fig:example-day}, where the daily course of
the clear sky reference irradiance and the CE and ECE identification
thresholds are shown along with the actual GHI measurements. In
addition, we provide an example scatter plot between the measured and
the modeled clear-sky irradiance for one year, where the CE and ECE
events are clearly grouped above the threshold of irradiance
(Figure\nobreakspace{}\ref{fig:example-year}).

@@

Following the application of the above discussed methodology to the
entire dataset (6.1 million one-minute GHI measurements), \(1.764\,\%\)
were identified as CE events and \(0.036\,\%\) as ECE events. The
highest recorded GHI due to CE was \(1416.6\,\text{W}/\text{m}^2\) on
24~May 2007, corresponding to OIR of \(345.9\,\text{W}/\text{m}^2\). The
stronger CE event of \(49.7\,\%\) above the clear sky threshold was
observed on 28~October 2016. In the following sections we are discussing
the long-term trends and variability of the CE events as well as of the
corresponding OIR and excess irradiation .

@@

We computed the daily trend of the mean OIR of CE using a first-order
autoregressive model with a lag of one day, with the `maximum
likelihood' fitting method \citep{Gardner1980, Jones1980} by
implementing the function `arima' from the library `stats' of the R
programming language \citep{RCT2023}. The trends are reported together
with the \(2\sigma\) error. We observed a negligible change of
\(0.087\pm 0.08\,\text{W}/\text{m}^2/\text{year}\) on the mean OIR.
Although, the previous result is closer to the raw data, we preferred to
present the annual statistics, that give a more clear picture about the
long term CE trends. Hence, we calculated the annual values from the
one-minute measurements. Despite the very weak statistical significance
of the mean OIR trend, the annual number of CE occurrences shows a
steady increase of \(45.6\pm 21.9\,\text{cases}/\text{year}\)
(Figure\nobreakspace{}\ref{fig:P-energy-N}). Subsequently, the mean
annual energy of the CE events is increasing with a rate of
\(116.9\pm 67.8\,\text{kj}/\text{year}\)
(Figure~\ref{fig:P-energy-sum}), witch follows the trend of the number
of CE occurrences. We have to note that the energy related to the CE
events can not be directly linked with the total energy balance of the
atmosphere. The net solar radiation of the region is not increased, but
rather redistributed through the CE events. This also is depicted by the
``instantaneous'' values of irradiance (average over one minute), that
can exceed the equivalent clear sky irradiance to a considerable level.

\begin{figure}[h!]
        {\centering 
            \subfloat[All skies.\label{fig:cusumnotrendmonthly-1}]
                {\includegraphics[width=\linewidth]{../images/P-energy-3} }\\
            \subfloat[Clear skies.\label{fig:cusumnotrendmonthly-2}]
                {\includegraphics[width=\linewidth]{../images/P-energy-2} }\\
            \subfloat[Cloudy skies.\label{fig:cusumnotrendmonthly-3}]
                {\includegraphics[width=\linewidth]{../images/P-energy-1} }
        }
    \caption{Time series for the period 1992 -- 2023 of (a) the yearly CE number of occurrences, (b) the yearly mean OIR and (c) the yearly excess irradiation. The black lines represent the linear trends on the yearly data.}\label{fig:cusumnotrendmonthly}
\end{figure}

@@

\hypertarget{appendix}{%
\section*{Appendix}\label{appendix}}
\addcontentsline{toc}{section}{Appendix}

>>>>

According to this study, in the period 1997 -- 2005 the mean AOD at
\(340\,\text{nm}\) \(0.403\) with a trend of \(-3.8\pm0.93\,\%\) per
year, corresponding to a change of \(0.0153\) per year. Using an
Ångström coefficient \(\alpha = 1.6\), this translates to a change in
the Ångström coefficient \(\beta=0.00272\) per year (or \(\beta=0.084\)
in 1997 and \(\beta=0.059\) in 2005). Simulations with uvspec for the
above Ångström coefficients, with WC of \(15.7\,\text{mm}\) and
\(15.4\,\text{mm}\) for 1997 and 2005 respectively and for a SZA of
\(55^\circ\) reveal a trend of \(+0.21\,\%\) per year in
\(\text{GHI}_\text{ref}\). The SZA of \(55^\circ\) was chosen as
representative of all days in the year in order to get a rough estimate
of the annually averaged change in clear sky irradiance. For the period
2005 -- 2023 we used the mean monthly values of AOD and WC from AERONET
in a similar simulation scheme to calculate the monthly mean of clear
sky irradiance, and finally the trend of \(+0.14\,\%\) per year. We
applied these two long term changes (see
Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
\(\text{GHI}_\text{ref}\), in order to create a more realistic
representation of the clear-sky irradiance for the whole period of
study.

@@

First, we evaluated the performance of the modelled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmosphere in Thessaloniki with lower
than average load of aerosols and humidity, which are the main factor
that attenuate the GHI, excluding clouds. With this approach the
simulated \(\text{GHI}_\text{ref}\) should be generally greater than the
measured GHI when aerosols are more abundant. To compensate for this, we
increased the \(\text{GHI}_\text{ref}\) by (\(4\,\%\)) with an
additional constant offset of \(15\,\text{W}/\text{m}^2\), as described
in Equation\nobreakspace\ref{eq:CE4}. \begin{equation}
\text{CE} : \text{E}_\text{i} > 15 + 1.04 \cdot \text{E}_\text{CSm,i} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation} where: \(\text{E}\) the measured irradiance,
\(\text{E}_\text{CSm}\) the selected modelled clear sky irradiance, and
\(i\) each of the one-minute observation. This is the criterion of our
CE identification.

@@

The definition of the CE events with this method has a degree of
subjectivity, since the actual clear sky irradiance is not known and can
only be approximated. However, this method is capable in detecting all
major CE events. Where some CE events with very low OIR may be not
detected, these are few with small over-irradiance and it is unlikely
that will effect significantly our results.

A sub-category of the CE events that is often discussed in the relevant
bibliography \citep{Cordero2023, Martins2022, Yordanov2015}, are the
extreme cloud enhancement (ECE) events. These are cases of CE where the
measured intensity of the irradiance exceeds the TSI at the top of the
atmosphere: \begin{equation}
\text{ECE}: \text{GHI}_\text{i} > \cos(\theta) \times E_{i\odot} / r_{i}^2
\label{eq:ECE}
\end{equation} where: \(\theta\) the solar zenith angle, \(E_{\odot}\)
the solar constant adjusted for the actual Sun -- Earth distance, and
\(i\) each of the one-minute observation.

An example of CE identification for a selected day is given in the
Figure\nobreakspace{}\ref{fig:example-day}, where the daily course of
the clear sky reference irradiance and the CE and ECE identification
thresholds are shown along with the actual GHI measurements. In
addition, we provide an example scatter plot between the measured and
the modeled clear-sky irradiance for one year, where the CE and ECE
events are clearly grouped above the threshold of irradiance
(Figure\nobreakspace{}\ref{fig:example-year}).

@@

Following the application of the above discussed methodology to the
entire dataset (6.1 million of one-minute GHI measurements),
\(1.764\,\%\) were identified as CE events and \(0.036\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007, corresponding to OIR of
\(345.9\,\text{W}/\text{m}^2\). The stronger CE event of \(49.7\,\%\)
above the clear sky threshold was observed on 28~October 2016. In the
following sections we are discussing the long-term trends and
variability of the CE events as well as of the corresponding OIR and
excess irradiation .

@@

The main focus of this study is to investigate the time evolution of the
CE events by analyzing the GHI measurements at Thessaloniki. Cloud
enhancements can be influenced by different factors, such as the
geometry, size and optical thickness of clouds, their height in the
atmosphere and on local weather regimes \textbf{!!!(ref)!!!}. Some of
these factors are related to changes in climate; hence it would be
reasonable to expect capturing their contributions to the frequency of
occurrence of CE events over Thessaloniki, as well as to the average OIR
and excess irradiation. The long-term trends were calculated using a
first-order autoregressive model with the `maximum likelihood' fitting
method \citep{Gardner1980, Jones1980}, by implementing the function
`arima' from the library `stats' of the R programming language
\citep{RCT2023}. All trends are reported together with their \(2\sigma\)
error.

We observed a negligible change of
\(0.087\pm 0.08\,\text{W}/\text{m}^2/\text{year}\) on the mean OIR.
Although, the previous result is closer to the raw data, we preferred to
present the annual statistics, that give a more clear picture about the
long term CE trends. Hence, we calculated the annual values from the
one-minute measurements. Despite the very weak statistical significance
of the mean OIR trend, the annual number of CE occurrences shows a
steady increase of \(45.6\pm 21.9\,\text{cases}/\text{year}\)
(Figure\nobreakspace{}\ref{fig:P-energy-N}). Subsequently, the mean
annual energy of the CE events is increasing with a rate of
\(116.9\pm 67.8\,\text{kj}/\text{year}\)
(Figure~\ref{fig:P-energy-sum}), witch follows the trend of the number
of CE occurrences. We have to note that the energy related to the CE
events can not be directly linked with the total energy balance of the
atmosphere. The net solar radiation of the region is not increased, but
rather redistributed through the CE events. This also is depicted by the
``instantaneous'' values of irradiance (average over one minute), that
can exceed the equivalent clear sky irradiance to a considerable level.

\begin{figure}% [h!]
        {\centering 
            \subfloat[\label{fig:cusumnotrendmonthly-1}]
                {\includegraphics[width=\linewidth]{../images/P-energy-3} }\\
            \subfloat[\label{fig:P-energy-N}]
                {\includegraphics[width=\linewidth]{../images/P-energy-2} }\\
            \subfloat[\label{fig:cusumnotrendmonthly-3}]
                {\includegraphics[width=\linewidth]{../images/P-energy-1} }
        }
    \caption{Time series for the period 1992 -- 2023 of (a) the yearly CE number of occurrences, (b) the yearly mean OIR and (c) the yearly excess irradiation. The black lines represent the linear trends on the yearly data.}\label{fig:cusumnotrendmonthly}
\end{figure}

@@
